---
title: "Linear Regression With a Categorical Exposure Variable"
author: "Abdirasak"
date: "2024-11-07"
output: 
  html_document:
    code_download: true 
    code_folding: show
    toc: true 
    toc_float:                          
      collapsed: false                  
      smooth_scroll: true               
    theme: readable 
    highlight: zenburn
    number_sections: true
    df_print: paged
    fig_caption: true
---
# Lung transplant data

For lung transplantation it is ideal for the donor’s lungs to be of a similar size to those of the recipient. Total lung capacity (TLC) is difficult to measure, so it is useful to be able to predict TLC from other information. These data are pre-transplantation TLC of 32 recipients of heart-lung transplants, obtained by whole-body plethysmography.

Variables: 
 - Age : Measured in years
 - sex : 1 = male; 2 = female
 - Height : measured in cm
 - TLC : Total lung capacity (litres)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library("tidyverse")
lung <- read.csv("lung.csv")
table(View(head(lung)))
str(lung)

#Descriptive statistics
summary(lung) # simple descriptive statistic
table(lung$sex)
hist(lung$age, main = "Age Distribution", xlab = "Age")

hist(lung$height, main = "Height Distribution", xlab = "Height(cm)")

hist(lung$tlc, main = "TLC", xlab = "TLC (litres)")
```

#Formatting data

```{r}
table(lung$sex) #the variable sex is currently stored as 1, and 2
str(lung$sex) #the sex variable is currently stored as an integer

lung$sexfactor <- factor(lung$sex, levels = c(1,2), labels = c("male", "female"))

head(lung) # sex is now a categorical variable
table(lung$sexfactor) # the label seems just about right.

boxplot(tlc~sexfactor, lung, xlab = "TLC and sex")
summary(lung$tlc)

# lets do some descriptive statistics

lung%>% 
  select(tlc, sexfactor) %>% 
  filter(sexfactor=="male") %>% 
  summary()

lung%>% 
  select(tlc, sexfactor) %>% 
  filter(sexfactor=="female") %>% 
  summary()
  
```
## Special note

I am gonna assume that the data doesnt meets the model assumptions


# Regression 

```{r}
simreg <- lm(tlc~sexfactor, lung)
summary(simreg)

lung$sexfactor2 <- relevel(lung$sexfactor, "female")
# The reference can be changed, if we need females to be the reference

simreg2 <- lm(tlc~sexfactor2, lung)
summary(simreg2)
# Out interpretation remains the same, regardless

```
# Interpretation 

The linear regression model shows that the average TLC for males is 6.9769, while females have an average TLC that is 1.7787 units lower, estimated at 5.1982. This difference is statistically significant (p = 0.000939). This particular  model explains 30.98% of the variation in TLC (adjusted R-squared = 0.2868). The overall model is statistically significant ( p-value of 0.0009392), indicating "sex" is a segnificant predictor of TLC.

# Multiple linear regression

```{r}
multreg <- lm(tlc~sexfactor+height,lung)
summary(multreg)
```
# Interpretation

The multiple regression model shows that height is also a strong positive predictor of TLC: each additional unit of height(cm) increases TLC by 0.07607 units, statistically significant (p = 0.00114). After adjusting for height, females have a TLC that is 0.77086 units lower than males, but this difference is not statistically significant (p = 0.13078). The model, in general, explains 52.37% of the variation in TLC, improving substantially over the sex-only model. The residuals are also reasonably small (residual standard error 1.158), indicating a better fit than the simpler model.


